Modelling fine-scale correlation structures using hidden Markov models

### 2020-07-17 Report/Sections/models.tex
Update Report

In this section we first remind the reader of the definition of a hidden Markov model (HMM). We then review three variations on the HMM: the CarHMM, HHMM, and HMM-DFT. We also briefly describe a continuous-time model and its connection with the CarHMM. Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The hidden Markov model (HMM)}

A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i$th component
$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
%
Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
%
(Fig \ref{fig:HMM}) represents the dependence structure of an HMM.

To find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$, let $y = (y_1,\ldots,y_T)$ be the observed emissions. 
We write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
%
$$\calL_{\text{HMM}}(y;\Theta,\Gamma) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry  $f^{(i)}(y_t; \theta^{(i)})$.
%

We will always parameterize transition probability matrices to remove the constraints that the entries of the matrix are non-negative and that the rows sum to 1.  We do this by parameterizing the $N \times N$ transition probability matrix $\Gamma$ using $\eta \in \bbR^{N \times N}$ \citep{Barajas:2017}:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
setting $\eta_{ii}$  to zero, $i=1,\ldots, N$, for identifiability.  Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using any numerical optimizer.  For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The conditionally auto-regressive hidden Markov model (CarHMM)}

A key assumption of an HMM is \textit{conditional independence} between observations given the state sequence.  Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.

The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicitly models auto-correlation within an HMM. Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$. 
The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
As an example, we could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%
\subsection{Continuous-time processes}

While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals. Another way to model a stochastic process is to use a stochastic differential equation (SDE). As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. This SDE is referred to as a state-switching Ornstein-Uhlenbeck process. Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit. However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If:
\begin{enumerate}
    \item The hidden behavioural process $X$ from (eq. \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
    \item Observations of the SDE from (eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

\subsection{The hierarchical hidden Markov model (HHMM)}

A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs. The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.   
%
In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$. The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner. Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$. Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$. The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study in section \ref{sec:case_study}, we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$). Figure \ref{fig:HHMM} shows a graphical representation of the dependence structure for an HHMM.

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^*$ be the $(T_1^* + \cdots + T_T^*)$-vector of the observed fine-scale emissions.
%
Let $\Theta^*$ denote the collection of all fine-scale emission parameters,
$\Theta^{*(i)}$, $i=1,\ldots, N$, and let $\Gamma^*$ denote the collection of all fine-scale transition probability matrices, $\Gamma^{*(i)}$, $i=1,\ldots, N$.
%
The likelihood of the observed data is
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citep{Adam:2019}.

@@

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity, we assume that the length of the fine-scale processes is constant (i.e. that $T^*_t = T^*$), although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour. We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.

Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

It is possible to accommodate for unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT). We do not describe this method here, but the generalization is straightforward. Refer to \citep{Bagchi:1999} for details.

@@

In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textit{any} parametric model which admits an easy-to-compute likelihood. The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$. This definition is straightforward to extend to the CarHMM as well:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.

>>>>

Traditional hidden Markov models (HMMs) are useful tools to \textit{blah blah blah - quick (5-7words) restatement of the main advantage of HMM}, and will be used as the core structure of our models.
However, traditional HMMs assume conditional independence between observations given the state sequence, and thus do not hold when the observations exhibit certain forms of significant correlation in time. Traditional HMMs model process autocorrelation with a single Markov chain, however many processes have temporal dependencies acting at different scales. \textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}. 
To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM-DFT. We will show how each of these variations can be altered and combined to form a wide variety of new models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The base structure of HMMs}

An HMM is comprised of a sequence of unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$.
The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
%
Figure \ref{fig:HMM} represents the dependence structure of an HMM.

Using observed emissions, here denoted $y = (y_1,\ldots,y_T)$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$. We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(y;\Theta,\Gamma) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
%

Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ in a way that forces the entries of the matrix to be non-negative and for the rows to sum to 1:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.  Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizer.  For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relaxing the conditional independence assumption with a CarHMM}

The CarHMM, introduced by Lawer et al. (2019), explicitly models auto-correlation in observation within an HMM. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$. 
The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. Figure \ref{fig:CarHMM} shows the dependence structure of a CarHMM.

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
For example, one could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.

While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals. Another way to model a stochastic process is to use a stochastic differential equation (SDE). As an example, Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit. However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If the following conditions are met:
\begin{enumerate}
    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,

    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

\subsection{Incorporating different scales of correlations with HHMM}

HHMM accounts for different levels of correlation by modeling a coarse-scale process and a fine-scale process, each of which with an HMM \textit{CITATION(s)}. The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.   
%
In the hierarchical setting, each state $X_t$ emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. The fine-scale process ($X_t^*, Y_t^*$) makes up another HMM with parameters that depend on the value of $X_t$. Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.

The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.We denote the fine-scale emission parameter vector corresponding to $X_t=i$ with $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$.

Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$). Figure \ref{fig:HHMM} represents the dependence structure for an HHMM.

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^*$ be the $(T_1^* + \cdots + T_T^*)$-vector of the observed fine-scale emissions.
%
Let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and let $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see Adam et al. (2019).

@@

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity, we assume that the length of the fine-scale processes is constant (i.e. that $T^*_t = T^*$), although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour. We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right), \quad k = 0, 1, \ldots, T^*-1.
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
In words, $z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure \ref{fig:HMM-DFT} shows a graphical representation of the HMM-DFT.

Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT) \citep{Bagchi:1999}. We do not describe the method in detail here, but the generalization is straightforward.

@@

In addition to the models described above, the fine-scale process $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood. The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the general fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$. This definition is straightforward to extend to the CarHMM as well:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
